Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-08-12 treestats_paper.tex
node timing plot script
node f4

%%%%%%%%%%%%%%%%%%
\subsection*{Empirical duality between branch and site statistics?}

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them.
To evaluate the correspondence between site and branch statistics
in real data,
we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
in 1Mb windows separately in each of the five continental groupings,
calculated using the site statistic described in Example~\ref{ex:site_diversity},
and compared this to the corresponding branch statistic (of Example~\ref{ex:branch_diversity})
as calculated from three inferred tree sequences:
by tsinfer \citep{tsinfer}, which does not attenpt to estimate well-calibrated node times,
(2) by GEVA \citep{geva}, which aims to calibrate the node times reported by tsinfer,
and (3) by Relate \citep{relate}, an independently developed method.
Separate curves for each of the 26 populations are shown in Supplemental Figure \ref{fig:chr20_full}.

@@

>>>>

%%%%%%%%%%%%%%%%%%
\section*{Application to 1000 Genomes tree sequences}

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them.
To evaluate the correspondence between site and branch statistics
in real data,
we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
in 1Mb windows separately in each of the five continental groupings,
calculated using the site statistic described in Example~\ref{ex:site_diversity},
and compared this to the corresponding branch statistic (of Example~\ref{ex:branch_diversity})
as calculated from three inferred tree sequences:
by tsinfer \citep{tsinfer}, which does not attenpt to estimate well-calibrated node times,
(2) by GEVA \citep{geva}, which aims to calibrate the node times reported by tsinfer,
and (3) by Relate \citep{relate}, an independently developed method.
All calculations were done only using the portions of the chromosome
passing the ``strict'' mask for sequencing accessibility.
Separate curves for each of the 26 populations are shown in Supplemental Figure \ref{fig:chr20_full}.

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Time stratification of statistics}

Thus far we have examined how statistics vary along the genome.
A tree sequence allows us access to another dimension: time.
Next, we use node statistics to display the \emph{time} distribution of admixture signal
in the tree sequences inferred by tsinfer/GEVA and by Relate.
The ``node'' $f_4(A, B; C, D)$ statistic for four populations $A$, $B$, $C$, and $D$
can be used to calculate, for each node, the average of $(p_A - p_B)(p_C - p_D)$
across the portion of the genome that the node is present in the tree sequence,
where $p_A$ is the proportion of samples in $A$ that lie below that node, etcetera.
A node with a large value will contribute a large amount to the site or branch $f_4$.

\begin{figure}
    \includegraphics{{figures/1kg/relate_chr20.node.f4}.pdf}
    \includegraphics{{figures/1kg/tgp_geva_chr20.node.f4}.pdf}
    \caption{
        Time distribution of signal for two $f_4$ statistics
        in tree sequences inferred by \textbf{(top)} Relate and \textbf{(bottom)} GEVA.
        Both 
        \label{fig:node_f4}
    }
\end{figure}
